Multi-resolution Neural Field Estimation

### 2011-03-06 Multiresolution_framework/ltx/manuscript.tex
changed EM equations

The MRAIDE can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate the convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)dr',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$ and $\ast$ denotes convolution and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$ ($rect$ function).
% \begin{equation}
% N_{1}\left(r\right)=
% \begin{cases}
% 1 & \text{if $ 0\le r<1$}, \\
% 0 & \mathrm{elsewhere}.
% \end{cases}
% \end{equation}
Equation \eqref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ convolutions of the indicator function with itself since
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is the inner product between two B-splines is another translated B-spline, such that
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. This holds as $N_m\left(r\right)$ is isotropic. The $m^{th}$-order B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} \frac{\left(-1\right)^n}{2^{m-1}}  q_n N_{m}\left(2r-n\right) \\
 q & _n = \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct} the integrals in \eqref{eq:Lambdax} and \eqref{eq:Lambdatheta} can be computed analytically. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale. In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in \eqref{eq:Lambdax} and \eqref{eq:Lambdatheta}.

% \cut{ It is important to note that B-spline scaling and wavelet functions have the following orthogonality properties \cite{Unser1993}: 
% \begin{equation}
%  \left\langle \psi_{j_1,l_1}(s),\psi_{j_2,l_2}(s)\right\rangle =0  \quad \mathrm{for} \quad j_1\neq j_2
% \label{PsiPsiOrthogonality}
% \end{equation}
% \begin{equation}
%  \left\langle \phi_{j_1,l_1}(s),\psi_{j_2,l_2}(s)\right\rangle =0  \quad \mathrm{for} \quad j_1\leq j_2.
% \label{PhiPsiOrthogonality}
% \end{equation}
\section{State and Parameter Estimation}
The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For the linear state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound used is a quadratic of the form
\begin{equation}\label{eq:Qfunction}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\varsigma\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\varsigma\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
where $\beta$ is constant with respect to $\boldsymbol\theta$. Equation \eqref{eq:Qfunction} is maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0 & =\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}\left[ \sum_{k=1}^{n_x}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{j,k}[\mathbf U]^{k,i}\right] \\ 
 \boldsymbol\upsilon_1 & =\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}\left[\sum_{k=1}^{n_x}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{j,k}[\mathbf U]^{k,i}\right] \label{eq:upsilon1}\\
 \boldsymbol\Upsilon&=T_s\varsigma\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}\left[\sum_{k,m=1}^{n_x}[\mathbf{U}^{\top}]^{i,m}[ \boldsymbol\Lambda_{x}^{-1}\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{m,k}[\mathbf{U}]^{k,j}\right]\label{eq:Upsilon}.
\end{align}
where $[.]_{i,j}$ denotes $\left(i,j\right)$-element of the matrix, and $ [\mathbf U]^{i,j}$ denotes $\left(i,j\right)$-block of the $n_x \times n_x$ block matrix $\mathbf U$. Each $1\times n_{\theta}$ block of $\mathbf U$ is given by
\begin{align}
\left[ \mathbf U\right] ^{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(r-r'\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. All $\left(i,j\right)$-blocks in the second brackets in  (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed as a once-off before the commencement of the EM iterations. Note that to separate $\boldsymbol\theta$ from $\boldsymbol\Lambda_{\theta}$ in forming the $\mathcal{Q}$-function we used the relation
\begin{equation}
\left[ \boldsymbol\Lambda_{\theta}\right] _{i,j} =\left[ \mathbf U\right]^{i,j}\boldsymbol\theta
\end{equation}

>>>>

The MRAIDE can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate the convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)dr',
\label{SplineConvolutionIntegral}
where $m>1$ and $\ast$ denotes convolution and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$ ($rect$ function).
% \begin{equation}
% N_{1}\left(r\right)=
% \begin{cases}
% 1 & \text{if $ 0\le r<1$}, \\
% 0 & \mathrm{elsewhere}.
% \end{cases}
% \end{equation}
Equation \eqref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ convolutions of the indicator function with itself since
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is the inner product between two B-splines is another translated B-spline, such that
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. This holds as $N_m\left(r\right)$ is isotropic. The $m^{th}$-order B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} \frac{\left(-1\right)^n}{2^{m-1}}  q_n N_{m}\left(2r-n\right) \\
 q & _n = \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct} the integrals in \eqref{eq:Lambdax} and \eqref{eq:Lambdatheta} can be computed analytically. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale. In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in \eqref{eq:Lambdax} and \eqref{eq:Lambdatheta}.
\section{State and Parameter Estimation}
The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For the linear state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound used is a quadratic of the form
\begin{equation}\label{eq:Qfunction}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\varsigma\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\varsigma\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
where $\beta$ is constant with respect to $\boldsymbol\theta$. Equation \eqref{eq:Qfunction} is maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0 & =\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Gamma_1]^{j,i} \\ 
 \boldsymbol\upsilon_1 & =\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Gamma_2]^{j,i} \label{eq:upsilon1}\\
 \boldsymbol\Upsilon&=T_s\varsigma\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Gamma_3]^{j,i}\label{eq:Upsilon}.
\end{align}
where $[.]_{i,j}$ denotes $\left(i,j\right)$-element of the matrix, and $ [.]^{i,j}$ denotes $\left(i,j\right)$-block of the  block matrix and where
\begin{align}
 [\Gamma_1]^{j,i} & =\sum_{k=1}^{n_x}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{j,k}[\mathbf U]^{k,i}\label{eq:Gamma1} \\ 
 [\Gamma_2]^{j,i} & =\sum_{k=1}^{n_x}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{j,k}[\mathbf U]^{k,i}\label{eq:Gamma2} \\
[\Gamma_3]^{j,i}&=\sum_{k,m=1}^{n_x}[\mathbf{U}^{\top}]^{i,m}[ \boldsymbol\Lambda_{x}^{-1}\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}]_{m,k}[\mathbf{U}]^{k,j}\label{eq:Gamma3}.
\end{align}
 Each $1\times n_{\theta}$ block of $n_x \times n_x$ block matrix $\mathbf U$ is given by
\begin{align}
\left[ \mathbf U\right] ^{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(r-r'\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. All blocks of $\Gamma_1$,$\Gamma_2$ and $\Gamma_3$  can be computed as a once-off before the commencement of the EM iterations. Note that to separate $\boldsymbol\theta$ from $\boldsymbol\Lambda_{\theta}$ in forming the $\mathcal{Q}$-function we used the relation
\left[ \boldsymbol\Lambda_{\theta}\right] _{i,j} =\left[ \mathbf U\right]^{i,j}\boldsymbol\theta
